---
output:
  html_document:
    keep_md: yes
    theme: cerulean
---

Reproducible Research (Peer Assessment 1)
===========================================

Author: Nikos Alexandris, August 2014

## Loading the data

Reading the csv data file

```{r Load data, echo=TRUE}
activity <- read.csv("~/archive/coursera/Reproducible_Research/Peer_Assessment_1/activity.csv", stringsAsFactors=FALSE)
```  

Checking class & structure of imported data

```{r Class & Structure}
class(activity)
str(activity)
```

Summarising

```{r Summarise}
summary(activity)
```

Summarise data after *NAs omission*

```{r}
activity.nonas <- na.omit(activity)
summary(activity.nonas)
```

* * *

## What is mean total number of steps taken per day?

**1. Histogram of the total number of steps per day**

Firstly, we derive the total steps per day

```{r}
activity.aggregated.totals <- aggregate(activity$steps, list (activity$date), FUN = "sum")
colnames(activity.aggregated.totals) <- c ("Date", "Total")
```


Then, we may obtain the requested histogram via

```{r}
require(ggplot2)
qplot(activity.aggregated.totals$Total)
```

**2. Mean and Median of the total number of steps per day**

To derive descriptive parameters, we summarise the aggregated data,

```{r}
summary(activity.aggregated.totals)
```

The *mean* and *median* of the total number of steps per day, are **10766** and **10765** respectively as extracted from the ```summary()``` report shown above.


* * *

## What is the average daily activity pattern?

**1. Time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**

```{r}
require(plyr)
activity.nonas.intervals <- ddply (activity.nonas, .(interval), summarize, Total = sum (steps), Mean = mean (steps))
plot ( activity.nonas.intervals$interval, activity.nonas.intervals$Mean, type = "l", xlab = "Intervals", ylab = "Average Steps per Interval")
```

**2. The 5-minute interval, on average across all the days in the dataset, which contains the maximum number of steps,** can be retrieved via

```{r}
interval.max.row <- which.max (activity.nonas.intervals$Mean)
interval.max <- activity.nonas.intervals[interval.max.row, "interval"];
```

and is

```{r}
interval.max
```

A plot that depicts visually the above finding, can be obtained via the following instructions:

```{r Max-Interval}
plot ( activity.nonas.intervals$interval, activity.nonas.intervals$Mean, type = "l", xlab = "Intervals", ylab = "Average Steps per Interval", lwd=2, col = "gray")
abline(v=activity.nonas.intervals$interval[interval.max.row], col="red", xlab = "104", lty = 3, lwd=3)
axis(1, at=interval.max, col = "red", col.axis = "red")
```


* * *

## Imputing missing values

**1. Total number of missing values in the dataset**, i.e. the total number of rows with NAs, is **2304** as retrieved by ```summary(activity$steps)```

```{r, echo=FALSE}
summary(activity$steps)
```

**2. Filling missing values in the dataset** | *Simple approach, source: <http://www.mail-archive.com/r-help@r-project.org/msg58289.html>*

```{r}
impute <- function (x, fun) {
missing <- is.na(x)
replace ( x, missing, fun (x[!missing]))
}
activity.filled <- ddply(activity, ~interval, transform, steps = impute(steps, mean))
```

3. **Histogram, mean and median of the total number of steps per day**

As in *question 2*, we derive the total steps per day

```{r}
activity.filled.aggregated.totals <- aggregate(activity.filled$steps, list (activity.filled$date), FUN = "sum")
colnames(activity.filled.aggregated.totals) <- c ("Date", "Total")
```


Then, we may obtain the requested histogram via

```{r}
require(ggplot2)
qplot(activity.filled.aggregated.totals$Total)
```

Next, we derive the *mean* and *median* values of the total number of steps per day

```{r}
require(plyr)
activity.filled.ddply <- ddply(activity.filled, .(date), summarize, Totals=sum(steps), Mean=mean(steps))
summary(activity.filled.ddply)
```
From the ```summary()``` tables shown above, the requested values are identical and equal to **10766**.

**Original vs. Filled-in Data**

The *mean* value is identical to the one of the original data set. The *median* has been equalised to the *mean* and differs only slightly from the original one (i.e. now **10766** vs. **10765**).  The impact of imputing missing data on the estimates of the total daily number of steps, using a very simple (simplistic?) approach, is minimal as the distribution has not been altered strongly.  

The effect of this simple fill-in approach, can be seen in a comparison box-plot of the *original* total number of steps per day (including missing observations) versus the *filled* values.

First we combine the vectors of interest in a new data.frame,

```{r}
original.vs.filled <- cbind(activity.aggregated.totals, activity.filled.aggregated.totals$Total)
colnames(original.vs.filled) <- c ("Date", "Original", "Filled")
```

and then we can box-plot them all in once step:
```{r}
boxplot(original.vs.filled[,-1], main = "Original vs. Filled", sub = "Total Number of Steps per Day", col = "green")
```
* * *

## Are there differences in activity patterns between weekdays and weekends?

**1. New factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day**

```{r}
activity.filled$Day <- format ( as.POSIXct ( activity.filled$date), "%A" )
activity.filled[!activity.filled$Day %in% c('Saturday', 'Sunday'),]$Day <- 'weekday'
activity.filled[activity.filled$Day %in% c('Saturday', 'Sunday'),]$Day <- 'weekend'
activity.filled$Day <-as.factor(activity.filled$Day)
```

**2. A panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).**

```{r}
require(lattice)
xyplot(steps ~ interval | Day, data=activity.filled, layout = c (1,2), panel = function(x, y, ...) {
  panel.xyplot(x, y, type = "l")
  panel.abline(h = mean(y), lty = 2, col = "red")
})
```
